---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(magrittr)
```

# goodstart (0.0.0.9006)

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![R build status](https://github.com/rappster/goodstart/workflows/R-CMD-check/badge.svg)](https://github.com/rappster/goodstart/actions)
[![Codecov test coverage](https://codecov.io/gh/rappster/goodstart/branch/master/graph/badge.svg)](https://codecov.io/gh/rappster/goodstart?branch=master)
<!-- badges: end -->

Have a good start when developing packages

## What?

`{goodstart}` tries to ensure a number of workflow aspects that are considered
best practice for R developers to provide you with an initial package state
where everything is "all set and ready to go".

It does so by providing convenience wrappers (function family `ensure_*()`)
around best practice functionality of [{usethis}](https://usethis.r-lib.org/) &
friends - mixed with some custom functions for automating some of the stuff I
usually do when starting a new development project.

The goal is basically to give you - the R developer - a "one-stop shop" function
that bundles all those calls to `usethis::use_*()` functions and associated
setup steps that you would normally perform when setting up a new package
project.

```r
goodstart::ensure_good_start()
```

## Why?

Reduce the ramp up time!

Even with the great tooling around [{usethis}](https://usethis.r-lib.org/) &
friends, setting up and configuring each new package project usually still takes
me about 30 to 60 minutes until I feel like "ok, now I'm ready to actually go
and write the first line of code of what the package will be all about". 

That's way too long.

For me, the main reason it takes so long is actually **remembering** all the
steps I have to do - all the cool `usethis::use_*()` functions I have to call,
all my other packages where "I used something that I'd like to use again", all
the nitty gritty details of setting up config files to use things like [Travis
CI](https://travis-ci.org/), [GitHub
Actions](https://github.com/features/actions) or [Codecov](https://codecov.io/),
etc.

`{goodstart}` tries to be your "one-stop shop" for dealing with all of that for you.

## Disclaimer

I started developing this package to scratch my own itch. 

Thus, while do I try to respect and follow as many of the emerging best
practices, the package in **its current state is very opinionated** to
streamline my own development process. It's also still very experimental and
still has a lot of rough edges (see my approach with `ask()` rounding these off
eventually).

**However, I'll try my best to adapt it to the needs of other developers as I go
along in case there's interested.**

## Acknowledgments & Pledge

I'm **absolutely** fascinated by all the great workflow tooling that has emerged
and continues to do so for R developers and huge thanks go out to all the
developers and teams that create and maintain it.

Just a couple of awesome examples:

- [{usethis}](https://usethis.r-lib.org/)
- [{devtools}](https://devtools.r-lib.org/)
- [{goodpractice}](http://mangothecat.github.io/goodpractice/)
- [{lintr}](https://www.tidyverse.org/blog/2017/12/workflow-vs-script/)
- [{testthat}](https://testthat.r-lib.org/)
- [{tinytest}](https://github.com/markvanderloo/tinytest)
- [{covr}](http://covr.r-lib.org/index.html)
- [{assertr}](https://github.com/ropensci/assertr)
- [{daff}](http://paulfitz.github.io/daff/)
- [{lumberjack}](https://github.com/markvanderloo/lumberjack)
- [{precommit}](https://lorenzwalthert.github.io/precommit/)
- [{drake}](https://docs.ropensci.org/drake/index.html)
- https://happygitwithr.com/ 
- Blog posts & Co.
    - [R Hub: Workflow automation tools for package developers](https://blog.r-hub.io/2020/04/29/maintenance/)
    - [R Hub: Why and how maintain a NEWS file for your R package?](https://blog.r-hub.io/2020/05/08/pkg-news/)
    - [Tidyverse: Self-cleaning test fixtures](https://www.tidyverse.org/blog/2020/04/self-cleaning-test-fixtures/)
    - [Miles McBain: Benefits of a function-based diet](https://milesmcbain.xyz/the-drake-post/)
    - [Mark van der Loo: Lumberjack JSS paper](https://cran.r-project.org/web/packages/lumberjack/vignettes/JSS_4008.pdf)
- ...

While I'm not yet as familiar as I'd like to be with all of the above paradigms,
workflows and packages, I'll try follow and support emerging best practices and
(de facto) standard workflows the best that I can as this package evolves.

As Microsoft CEO Satya Nadella has once put it:

> If any engineer has to choose between working on a feature or working on dev
productivity, always choose dev productivity.

So in that spirit, I hope that `{goodstart}` can eventually contribute its 2
cents in the quest of establishing clean and reproducible development workflows
in R.

Any suggestions or help along the way are greatly appreciated.

## Installation

### Stable releases on [CRAN](https://CRAN.R-project.org)

Not available yet.

### Stable releases on [GitHub](https://github.com/)

``` r
remotes::install_github(
  "rappster/goodstart",
  ref = "release-{release_version}"
)
```

This installs the version in Git branch `release-{release_version}`

Substitute placeholder `{release_version}` with a [semantic version
number](https://semver.org/) of your choice (e.g. `0.0.1`).

Also make sure that the actual Git branch for your desired version exists by
inspecting the `branch` dropdown of this GitHub repository.

### Development releases on [GitHub](https://github.com/)

``` r
remotes::install_github("rappster/goodstart")
```

This installs the version in Git branch `master`

## Usage

The main function of `{goodstart}` is `ensure_good_start()`.

It bundles calls to a number of `ensure_*()` functions that mostly in turn call `usethis::use_*()` functions with some adde bells and whistles.

Prior to calling `ensure_good_start()`, I/you can decide about your config like
this:

```{r eval=TRUE}
config <- goodstart::goodstart_config(
  use_github_user_name = "rappster",
  use_github_repo_name = "pkgname",
  use_github_auth = "ssh"
)

config
```

Tweak that list as you see fit and then call

```{r eval=FALSE}
goodstart::ensure_good_start(config = config)
```

### Test drive

I actually put a lot of effort of making `{goodstart}` testable.

Oonce more: a shout out to the awesome developers behind
[{usethis}](https://usethis.r-lib.org/) and to [Jenny
Bryan](https://github.com/jennybc) in particular for her extremely insightful
and useful blog post on [Self-cleaning text
fixtures](https://www.tidyverse.org/blog/2020/04/self-cleaning-test-fixtures/) which helped **enormously** in writing unit tests. 

One nice side effect of the unit tests is that you can easily set up a temporary package "to play with":

```{r}
pkg <- goodstart::create_sandbox_package()

pkg
pkg %>% fs::dir_ls(all = TRUE)
```

As of now, you still need to do some "behind the scenes" prep stuff that I do
when testing:

```{r}
goodstart:::when_testing_load_local_package()
"R/hello.R" %>% usethis::proj_path() %>% fs::file_create() 
# Example file created when creating new package projects in RStudio
goodstart:::when_testing_ensure_man_dir()
"man/hello.Rd" %>% usethis::proj_path() %>% fs::file_create()
# Example file created when creating new package projects in RStudio
goodstart:::when_testing_namespace_default()
goodstart:::when_testing_ensure_example_function_and_unit_test()

goodstart:::ensure_removed_file("README.Rmd")
goodstart:::when_testing_copy_template("package-README")
goodstart:::when_testing_copy_template("package-BACKLOG")
```

But then you can safely call `ensure_good_start()`

```{r}
goodstart::ensure_good_start(config = config)
```

And inspect the results

```{r}
pkg %>% fs::dir_ls(all = TRUE)
```

Make sure to clean up things with

```{r}
withr::deferred_run()
```

to get back to your previous state (working directory, loaded projects, etc.)

### Be nice and ask

I also developed `ask()` as my first attempt to giving developers other than me
more transparency and choice regarding what's actually happening once you call
`ensure_good_start()`

It asks you interactively about the setup you'd like to use, either in "full verbosity" or reduced to the essentials (try `ask(minimal = TRUE)`) once you know what's going on.

Either call `ask()` or `ask(minimal = TRUE)`:

```{r eval=FALSE}
goodstart::ask()
goodstart::ask(minimal = TRUE)
```

This question/answer process results in a `config` object that has the same
structure as the one returned by `goodstart::goodstart_config()`.

This is how each step will look like:

### Step 1

```{r step-1, eval=FALSE}
ℹ Roxygen (step 1 of 10)

Great documentation makes a huge difference - not just for others, but also
for "future you".

Using package `roxygen2` (https://roxygen2.r-lib.org/) makes documenting
your package easy and fun. It offers you a straightforward syntax (optionally
including markdown), provides you with a number of convenient helpers and takes
care of managing your NAMESPACE file.

Answering Yes implies the following when subsequently calling `ensure_good_start()`:

✓ Package `roxygen2` will be installed if necessary
✓ Package `roxygen2` will be added to Suggests section of your
  DESCRIPTION file
✓ Line RoxygenNote: {x.y.z} will be added to your DESCRIPTION
  file
✓ [NOT ENSURED YET] RStudio settings are modified to put `roxygen2` in
  charge of documenting and vignette building

Do you want to use package `roxygen2` for documenting tasks? 

1: Yes
2: No
3: Let me start over
4: Exit
```

### Step 2

```{r step-2, eval=FALSE}
ℹ NEWS.md (step 2 of 10)

Using R Markdown (https://rmarkdown.rstudio.com/) makes your developer life
so much easier.

It offers you a very expressive syntax, makes it easy to write vignettes and
ties in excellently with `roxygen2` code (see https://roxygen2.r-
lib.org/articles/markdown.html).

Answering Yes implies the following when subsequently calling `ensure_good_start()`:

✓ Packages `rmarkdown`, `whisker` and `knitr` will be
  installed if necessary
✓ Packages `rmarkdown`, `whisker` and `knitr` will be
  added to the Suggests section of your DESCRIPTION file
✓ Line VignetteBuilder: knitr will be added to your DESCRIPTION
  file
✓ If you said Yes to `roxygen2` then line Roxygen:
  list(markdown = TRUE) will be added to your DESCRIPTION file

Do you want to use R Markdown throughout your package? 

1: Yes
2: No
3: Let me start over
4: Exit
```

### Step 3

```{r step-3, eval=FALSE}
ℹ License (step 3 of 10)

You should be using a license to make it explicit how you would like your
code to be used and shared.

Check out
  - http://r-pkgs.had.co.nz/description.html#license
  - https://www.r-project.org/Licenses/
for details on open source software licenses.

Answering Yes implies the following when subsequently calling `ensure_good_start()`:

✓ You are asked which license type you want to use. If you are not sure then
  GPL-3 is a good choice
✓ The line License: {license_type} will added to your
  DESCRIPTION file

Which license would you like to use? 

 1: GPL v3
 2: MIT
 3: CC0
 4: CCBY 4.0
 5: LGPL v3
 6: APL 2.0
 7: AGPL v3
 8: None
 9: Let me start over
10: Exit
```

### Step 4

```{r step-4, eval=FALSE}
ℹ Unit testing (step 4 of 10)

Unit testing is crucial if you're serious about continuous integration (CI)
as it gives you the confidence of iteratively improving code and adding new
features. If you break things, you'll immediately know and can triage where and
why things went wrong.

Package `testthat` offers you a sophisticated framework for unit testing
and plays nice with all major continuous integration (CI) platforms such as
  - https://travis-ci.org/
  - https://www.appveyor.com/
  - https://github.com/features/actions
via integration with https://codecov.io/ or https://coveralls.io/

Package `tinytest` is an alternative, but it is currently not further
integrated with `goodstart`.

Answering Yes implies the following when subsequently calling `ensure_good_start()`:

✓ The selected package will be installed if necessary
✓ Unit test infrastructure will be created within directory 'tests/'
✓ The selected package will be added to the Suggests section of your
  DESCRIPTION file

Which unit test package would you like to use? If you're not sure then
'testthat' is a good choice. 

1: testthat
2: tinytest
3: None
4: Let me start over
5: Exit
```

### Step 5

```{r step-5, eval=FALSE}
ℹ Unit testing: test coverage (step 5 of 10)

An important part of sincere continuous integration (CI) efforts is keeping
tabs on the proportion of your code that is tested via unit tests - or test
coverage in short.

Package `covr` provides an excellent framework for that. It also
integrates nicely with CI test coverage services such as
  - https://codecov.io/
  - https://coveralls.io/

Answering Yes implies the following when subsequently calling `ensure_good_start()`:

✓ Prerequisites:
  - You've said Yes to unit testing

Do you want to use a test coverage package? 

1: covr
2: None
3: Let me start over
4: Exit
```

### Step 6

```{r step-6, eval=FALSE}
ℹ Dependency management (step 6 of 10)

Systematically tracking tracking the exact state (i.e. version and where
it was installed from) of all the packages that your package depends on is
highly recommended. Using a dependency management framework greatly facilitates
colloboration as it lets others easily reproduce the exact same dependency state
when you share your code with them.

Using package `renv` (https://rstudio.github.io/renv/) is a great way of
managing your package dependencies.

Answering Yes implies the following when subsequently calling `ensure_good_start()`:

✓ The selected package will be installed if necessary
✓ The selected package will be added to the Suggests section of your
  DESCRIPTION file
✓ [RENV SPECIFIC] From now on, any package that you install will be installed
  to your package's local library at 'renv/library/R-{version}/
  {os_architecture>}'

Which dependency management package would you like to use? 

1: renv
2: None
3: Let me start over
4: Exit
```

### Step 7

```{r step-7, eval=FALSE}
ℹ GitHub (step 7 of 10)

GitHub (https://github.com) is a great way to collaborate and share
your code with others. Others can install your package from GitHub and also
systematically contribute to your package by reporting issues or even orovide
pull requests.

Answering Yes implies the following when subsequently calling `ensure_good_start()`:

✓ Prerequisites:
  - You have a GitHub account and know your account's user name
✓ [Next step] You'll be asked for your GitHub user name
✓ [Next step] You'll be asked for a GitHub repository name that will be
  set as your remote Git repository with name origin
✓ [Next step] You'll be asked how you want to authenticate to GitHub: via
  SSH (recommended) or via HTTPS authentication. Your answer
  determines the URL structure of the remote Git repository that is added:
  - SSH: 'ssh:/git@github.com:{user_name}/{repo_name}.git'
  - HTTPS: 'https:/github.com/{user_name}/{repo_name}.git'

Do you want to use GitHub? 

1: Yes
2: No
3: Let me start over
4: Exit
```

#### Step 7.1

```{r step-7.1, eval=FALSE}
What's your GitHub user name? Enter without quotes: rappster
```

#### Step 7.2

```{r step-7.2, eval=FALSE}
Is your GitHub repository name the same as your package's name ('goodstart')? 

1: Yes
2: No
3: Let me start over
4: Exit
```


#### Step 7.3

```{r step-7.3, eval=FALSE}
How would you like to authenticate to GitHub? 

1: ssh
2: https
3: Let me start over
4: Exit
```

### Step 8

```{r step-8, eval=FALSE}
ℹ Continuous integration (step 8 of 10)

Continuous integration (CI) is extremely important as it helps you deploy
early and often with the confidence of automatically checking your package on
different OS platforms (optionally also running your unit tests).

You can currently choose from the following CI platforms that support R
packages:
  - GitHub Actions (https://github.com/features/actions)
  - Travis CI (https://travis-ci.org/)
  - Appveyor (https://www.appveyor.com/)

Answering Yes implies the following when subsequently calling `ensure_good_start()`:

✓ Different side effects (creation of necessary artefacts such as configuration
  files, etc.) depending on your platform choice

Do you want to use a CI platform? 

1: github_actions
2: travis_ci
3: appveyor
4: Let me start over
5: Exit
```

### Step 9

```{r step-9, eval=FALSE}
ℹ Continuous integration: test coverage service (step 9 of 10)

If you use unit tests, quantify test coverage locally and use a continuous
integration platform, then it makes sense to include test coverage in your CI
pipeline

You can currently choose from the following CI platforms that are specialized on
test coverage and that support R packages:
  - Codecov (https://codecov.io/)
  - Coveralls (https://coveralls.io/)

Answering Yes implies the following when subsequently calling `ensure_good_start()`:

✓ Prerequisites:
  - You've said Yes to unit testing
  - You've said Yes to reporting test coverage
  - You've said Yes to using continuous integration (CI)
✓ Under the hood `use_coverage({type})` is called which triggers CI
  platform specific side effects (e.g. creating config YAML files, etc.)

Which CI test coverage service would you like to integrate with? 

1: codecov
2: coveralls
3: None
4: Let me start over
5: Exit
```

### Step 10

```{r step-10, eval=FALSE}
ℹ Code life cycle (step 10 of 10)

Using package `lifecycle` (https://lifecycle.r-lib.org/) helps
inform others about the life cycle state of your development

It provides a set of tools and conventions to manage the life cycle of your
exported functions and the package as a whole.

Answering Yes implies the following when subsequently calling `ensure_good_start()`:

✓ Package `lifecycle` will be installed if necessary
✓ Package `lifecycle` will be added to the Suggests section of
  your DESCRIPTION file
✓ Necessary infrastructure components such as graphics for badges etc. will be
  created within the 'man/' directory

Do you want to use package `lifecycle` to better inform users of your
package? 

1: Yes
2: No
3: Let me start over
4: Exit
```

## Future development

Going forward, amongst many other things, I'd like to:

1. Make it possible to persist and manage desired `config`s that either came out of
`goodstart_config()` or `ask()` in you user's home directory. 

    See `write_goodstart_config()`,  `read_goodstart_config()` and
    `exists_goodstart_config()`. Not nearly done, but it's a start.
    
2. Bundle all "intrusive actions" such as installing required packages, setting
up Git/GitHub stuff etc. in order to make them much more transparent, easier for
developers to assess and to customize to their needs.

3. Follow even more emerging best practices regarding paradigms, workflows and
actual tasks/steps/tools
